Supplementary materials

### 2018-07-11 WholePlotTrajectories.tex
pdf
biblio last reread

% User-adder preamble
\usepackage{textcomp} \DeclareUnicodeCharacter{B0}{\textdegree}
\usepackage{tabu} \usepackage{caption}
\captionsetup{justification = justified}
\renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
\hyphenation{bio-di-ver-si-ty sap-lings re-or-gan-i-za-tion}

@@

\Abstract{
Understanding the ecological rules underlying the maintenance of
tropical forests biodiversity, structure, functioning and dynamics is
urgent to anticipate their fate in the global change context. The huge
diversity of tropical forests is often assumed to be regularly reshaped
by natural disturbance yielding a diversity peak at intermediate
intensity. This intermediate disturbance hypothesis (IDH),though,
remains debated and the controversy questions the extent of communities'
resilience regarding their functional and taxonomic facets. To
disentangle the ecological processes driving community response to
disturbance, we analysed the diversity trajectories over 30 years
following a disturbance gradient. Specifically, we examined community
functional and taxonomic trajectories with regards to diversity,
composition and redundancy. Functional trajectories were drawn based on
7 leaf, stem and life-history traits. We highlighted the cyclic recovery
of community taxonomic and functional composition. While pre-disturbance
taxonomic differences were maintained over time, functional composition
trajectories were quite similar among communities. The IDH did predict
communities taxonomic diversity response while functional diversity was
enhanced whenever the disturbance intensity. Although consistent, the
recovery of community composition, diversity and redundancy remained
unachieved after 30 years. This acknowledged the need of decades-long
cycles with no disturbance to ensure a complete recovery, and questioned
tropical forest community resilience after repeated disturbance.
}

@@

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna,
regulate the local and regional climates, the carbon, water and nutrient
cycles, and ensure cultural and human well-being. The growing demand in
forests products together with current global changes increases the
pressure on remaining natural forests
\citep{Gibson2011a, Morales-Hidalgo2015} and threatens the maintenance
and dynamics in space and time of communities structure, composition and
functioning \citep{Anderson-Teixeira2013, Sist2015}.

In tropical forests, ecological communities are regularly re-shaped by
natural disturbance events changing both the abiotic environment,
through the fluxes of light, heat and water \citep{Goulamoussene2017},
and the biotic interactions such as competition among species
\citep{Chesson2000, Herault2018}. One of the cornerstone of tropical
forest ecology is to understand the processes and drivers of ecosystems
response to disturbance \citep{White2001, Chazdon2003a}. For now, this
has been largely studied through forest structural parameters such as
aboveground biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016} that are rapid and convenient to
measure. These structural parameters have been sucessfully modeled,
giving important insights into the recovery of ecosystem processes and
services \citep{Herault2018}. However the response of forests diversity
and composition is still unclear, albeit it determines the productivity,
stability and functioning of ecosystems \citep{Tilman2014, Liang2016}.
In the short-term, moderate disturbance may lead to positive impacts on
communities diversity, an idea formalized by the intermediate
disturbance hypothesis (IDH) stating a maximized species diversity when
disturbance intensity is not too high
\citep{Molino2001, Kariuki2006a, Berry2008a}.

Validations of the IDH though remain scarce in the long-term and mainly
rely on the analysis of taxonomic richness \citep{Molino2001}. Taxonomic
richness alone, though, gives limited or misleading information on
forests recovery and functioning \citep{Martin2015, Chaudhary2016}. More
ecological-meaningful analysis would couple richness with (i) evenness
that would reveal underlying ecological processes and (ii) composition
that is crucial for conservation issues
\citep{Magurran1988, Lavorel2002, Bellwood2006}. Furthermore, a
functional approach accounting for species biological attributes would
directly link communities diversity, composition and redundancy to
ecosystem functioning and to its environmental constraints
\citep{Violle2007b, Moretti2009, Baraloto2012a, Scheiter2013}. In that
respect, the functional trait-based approach that focus on major traits
related to species ecology and mediate species performance in a given
environment was sucessfully adopted \citep{Diaz2005, Villeger2008a}. For
instance, the functional approach revealed in tropical rainforests the
deterministic processes entailing, after disturbance, a functional shift
from a dominance of ``conservative'' slow-growing species dealing with
scarce resources to ``acquisitive'' fast-growing species with rapid and
efficient use of abundant resources
\citep{TerSteege2001, Reich2014, Herault2011}. This shift is translated
into the trajectories of key functional traits related to resource
acquisition (leaf and stem traits) and life-history traits (seed mass,
maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. Eventually
a complete overview of communities response to disturbance would
encompass the changes in functional redundancy, that quantifies the
amount of shared trait values among species \citep{Carmona2016}. The
high functional redundancy of hyperdiverse tropical forests
\citep{Bellwood2006} mitigates the impacts of species removal on
ecosystem functioning and determines the resilience of communities after
disturbance \citep{Trenbath1999, Elmqvist2003, Diaz2005}.

In this study, we monitored over 30 years the response of 75 ha of
neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of ecosystem above-ground biomass (AGB) loss. We made
use of a large functional traits database encompassing major leaf, stem
and life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy
\citep{Lohbeck2015, Guariguata2001}. Specifically, we (i) elucidated
community taxonomic and functional recovery and the determinism of
underlying processes, (ii) clarified the validity of the IDH in the long
term for tropical forest and its translation into different trajectories
in time, and (iii) questioned community recovery time.

@@

The experiment is a network of twelve 6.25ha plots that underwent a
disturbance gradient of three logging, thinning and fuelwood cutting
treatments (Table \ref{tab:Tab1}) according to a randomized plot design
with three replicate blocks of four plots. The disturbance corresponds
to averages of 10 trees removed per hectare with a diameter at 1.3 m
height (DBH) above 50 cm for treatment 1 (T1), 32 trees/ha above 40 cm
DBH for treatment 2 (T2) and 40 trees/ha above 40 cm DBH for treatment 3
(T3). Treatments T2 and T3 besides included the thinning of trees by
poison girdling \citep{Schmitt1990, Blanc2009}. The disturbance
intensity was measured as the percentage of aboveground biomass (\%AGB)
lost between the first inventory in 1984 and five years after
disturbance \citep{Piponiot2016} estimated with the BIOMASS R package
\citep{Biomass2018}.

@@

See appendix 1 and \citet{Aubry-Kientz2013} for the detailed
methodology.

@@

To counter taxonomic uncertainties due to the variability of botanical
identification levels (in space) and protocols (in time), the taxonomic
composition and diversity analysis were conducted at the genus level.
Taxonomic and functional trajectories of community composition were
followed in a two-dimensional NMDS ordination space. Two NMDS using
abundance-based (Bray-Curtis) dissimilarity measures were conducted to
map either taxonomic or functional composition, the later based on the 7
leaf, stem and life history traits (without seed mass classes).
Trajectories along time were reported through the euclidean distance
between the target inventories and the reference inventories in 1989,
\emph{i.e.} 2 years after disturbance. Univariate trajectories of the
leaf, stem and life-history traits were also visualized with the
community weighted means (CWM) \citep{Diaz2007, Garnier2004}. Species
seed mass corresponded to 5 classes of increasing mass, seed mass
trajectories were therefore reported as the proportion of each class in
the inventories (Supp Mat XX).

The taxonomic diversity was reported through species richness and
evenness, \emph{i.e} the Hill number translation of the Simpson index
\citep{Hill1973}. These indices belong to the set of HCDT or generalized
entropy, respectively corresponding to the 0 and 2 order of diversity
(q), recomended for diversity studies
\citep{Patil1982, Tothmeresz1995, Marcon2015}. The functional diversity
was reported using the functional richness and functional evenness,
\emph{i.e} Rao index of quadratic entropy which combines species
abundance distribution and average pairwise dissimilarity based on
species functional traits.

@@

The functional redundancy was measured as the overlap among species in
community functional space \citep{Carmona2016}. The samples of the trait
database were first mapped in a 2-dimensional plan with a PCA analysis.
Then, multivariate kernel density estimator associated with individual
trees returned species traits probability distribution (TDP). Species
TDP weighted by species abundance were eventually summed for each
community. Community functional redundancy was the sum of TDPs overlap,
expressed as the average number of species that could be removed from
without reducing the functional space (see appendix I for a more
comprehensive sheme).

@@

\caption{Plot trajectories in terms of flora composition (left panels \textbf{(a)} and \textbf{(c)}) and functional composition (right panels \textbf{(b)} and \textbf{(d)}) in a two-dimensional NMDS space. Lower panels (\textbf{(c)} and \textbf{(d)}) represent the euclidean distance to initial condition along the 30 sampled years. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals}\label{fig:NMDSplans}
\end{figure*}

@@

Maximum height at adult stage (\emph{Hmax}), leaf toughness
(\emph{L\_toughness}) and wood specific gravity (\emph{WSG}) first
decreased and then slightly increased but remained significantly lower
than their initial value (Figure \ref{fig:CWM}). On the other side, Bark
thickness (\emph{Bark\_thick}) and specific leaf area (\emph{SLA})
increased and while \emph{Bark\_thick} remained substantially high after
30 years, \emph{SLA} had almost recovered its initial value. For all
traits, the maximum difference to initial value was correlated to the
disturbance intensity (\(\rho_{spearman}^{L_{thickness}}=0.76\),
\(\rho_{spearman}^{L_{chloro}}=0.60\),
\(\rho_{spearman}^{L_{toughness}}=-0.53\),
\(\rho_{spearman}^{SLA}=0.93\), \(\rho_{spearman}^{WSG}=-0.75\),
\(\rho_{spearman}^{Bark-thickness}=0.71\),
\(\rho_{spearman}^{Hmax}=-0.40\)).

@@

\caption{Trajectories of the communities weighted means (CWM) over 30 years after disturbance of 4 leaf traits (Leaf thickness, \emph{L\_thickness}, chlorophyll content, \emph{L\_chloro}, toughness, \emph{L\_toughness} and specific area, \emph{SLA}), 2 stem traits (wood specific gravity, \emph{WSG}, and bark thickness, \emph{Bark-thick}) and one life history trait (Specific maximum height at adult stage, \emph{Hmax}). Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals.}\label{fig:CWM}
\end{figure*}

@@

\caption{Trajectories over 30 years of the difference with the 1989 inventory (2 years after disturbance) of community taxonomic \textbf{(a)} richness, \textbf{(b)}, taxonomic evenness, \textbf{(c)} functional richness, and \textbf{(d)} functional evenness. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals }\label{fig:DivTaxo}
\end{figure*}

@@

\caption{Relationship between the initial \%AGB loss and community taxonmic richness \textbf{(a)}, taxonomic evenness \textbf{(b)}, functional richness \textbf{(c)},and functional evenness \textbf{(d)} at 10, 20 and 30 years after disturbance. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals }\label{fig:IDHplot}
\end{figure*}

@@

\caption{Trajectories of the functional redundancy within the initial functional space over 30 years after disturbance. Colors are disturbance treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals.}(\#fig:RedFun_rest)
\end{figure}

@@

Conversely, disturbed communities followed functional trajectories
similar in terms of functional composition. As pre-disturbance surviving
trees mirror the initial community \citep{Herault2018}, changes in
functional composition relied upon the recruitment of species or
functional types that were infrequent or absent before disturbance.
Competitive pioneers became dominant in filling the environmental niches
of high availability of light, space and nutrient vacated by the
disturbance. The recruitment of pioneers changed community functional
composition in the same way for all disturbance intensity towards more
resource-acquisitive strategies, moving community functional composition
right along the first axis in Figure \ref{fig:NMDSplans}
\citep{Grime1998, Westoby1998, Wright2004, Reich2014}. Thereafter
long-lived, more resistant and shade-tolerant species excluded the first
established pioneers and started the recovery of pre-disturbance
functional composition, moving similarly community functional
composition left along the first axis and upward along the second axis
in Figure \ref{fig:NMDSplans} .

@@

The IDH predicted well the disturbance impact on community taxonomic
richness, enhanced until an intensity threshold (20-25\% AGB loss), and
to some extent on taxonomic evenness, somewhat decoupled from the
disturbance intensity as already observed in the Guiana Shield
\citep{Baraloto2012a} and in Bornean tropical forests
\citep{Cannon1998}. The disturbance intensity determined the balance in
the community between pre-disturbance surviving trees and those
recruited afterward. The pool of true pioneer species specifically
recruited after disturbance is restricted in the Guiana Shield to a few
common genera (e.g.~Cecropia spp., Vismia spp.)
\citep{Zalamea2012, Guitet2018}. Below the intensity threshold the size
of the surviving community maintained the pre-disturbance high taxonomic
richness while the recruitment of pioneers, infrequent or absent before
disturbance, increased both community taxonomic richness and evenness.
Beyond the intensity threshold, the disturbance decreased the taxonomic
richness of surviving trees which was not offset by the enrichment of
pioneers, so that the overall community taxonomic richness decreased
according to the disturbance intensity \citep{Molino2001}. For community
taxonomic evenness the disturbance impact was similar but slighter, as
the evenness is less sensitive to the loss of rare species. Taxonomic
evenness rather represented the increasing dominance of pionneers that
balanced the usual hyper-dominance of tropical forest below the
intensity threshold, thus increasing community overall evenness up to
the intensity threshold beyond which pioneers became in turn highly
dominant and decreased the overall evenness \citep{Baraloto2012a}.

@@

Along time, taxonomic richness trajectories of all disturbed communities
first dropped similarly, following the species loss due to disturbance,
and then displayed a species gain depending on the disturbance
intensity. Up to an intensity threshold, the species gain was all the
more significant that the disturbance intensity increased, with the
establishment of long-lived pioneers enhancing community taxonomic
richness and evenness in the long term. These long-lived pionneers,
functionally quite different from the functional composition, entailed
as well a progressive and long-lasting increase of the functional
richness and evenness \citep{Denslow1980, Molino2001}. Beyond an
intensity threshold, though, a few short-lived pioneers occupied the
vacated environmental space and prevented the establishment of other
species. These short-lived pioneers were functionally very different
from the pre-disturbance community and entailed a rapid and significant
increase of functional richness en evenness. Already after 10 years,
though, short-lived pioneers started to decline and the functional
richness and evenness decreased. Likely this decrease will be followed
by the establishment of long-lasting pioneers, and by the time they
recruit we expect the taxonomic and functional trajectories to catch up
with those observed after intermediate disturbance
\citep{Anderson2007, Walker2009}.

@@

For 15 years the species loss during disturbance, determined by the
disturbance intensity, commensurately decreased the functional
redundancy within the pre-disturbance functional space. The redundancy
decrease was not compensated in the first place as the first recruited
pionneers were functionally different from the pre-disturbance
functional composition. Progressively though, first established species
were replaced by more competitive long-lived pionneers or
late-successional species resembling more the pre-disturbance functional
composition and restoring the functional redundancy. This replacement
was stochastic and followed the lottery recruitment rules, implying a
recruitment eased for the first recruited species but then increasingly
hampered by the emergence of interspecific competition
\citep{Busing2002}. Along time the recovery of infrequent species was
increasingly difficult, so that the time for the full recovery of the
functional redundancy, in some communities just initiated after 30
years, was difficult to estimate
\citep{Trenbath1999, Elmqvist2003, Diaz2005}.

The long-term impact of disturbance on community functional redundancy
meant a lower resilience of the pre-disturbance communities, with higher
chances to see the persistence of disturbance-specific species at the
expense of late-successional ones
\citep{Haddad2008, Burslem2000, Martin2013}. Besides the long-term
recovery of infrequent species increases the risks to loose cornerstone
species, with unexpected ecological consequences
\citep{Jones1994, Chazdon2003a, Diaz2005, Gardner2007}. Apart from the
functional characteristics considered here, infrequent species might
indeed have unique functions in the ecosystem or be a key for some
fauna.

>>>>

% User-adder preamble
\usepackage{textcomp} \DeclareUnicodeCharacter{B0}{\textdegree}
\usepackage{tabu} \usepackage{caption}
\captionsetup{justification = justified}
\renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
\hyphenation{bio-di-ver-si-ty sap-lings re-or-gan-i-za-tion post-dis-tur-bance dis-tur-bance}

@@

\Abstract{
Understanding the ecological rules underlying the maintenance of
tropical forests biodiversity, structure, functioning and dynamics is
urgent to anticipate their fate in the global change context. The huge
diversity of tropical forests is often assumed to be regularly reshaped
by natural disturbance yielding a diversity peak at intermediate
intensity. This intermediate disturbance hypothesis (IDH), though,
remains debated and the controversy questions the extent of communities'
resilience regarding their functional and taxonomic facets. To
disentangle the ecological processes driving community response to
disturbance, we analysed the diversity trajectories over 30 years
following a disturbance gradient. Specifically, we examined community
functional and taxonomic trajectories with regards to diversity,
composition and redundancy. Functional trajectories were drawn based on
7 leaf, stem and life-history traits. We highlighted the cyclic recovery
of community taxonomic and functional composition. While pre-disturbance
taxonomic differences were maintained over time, functional composition
trajectories were quite similar among communities. The IDH did predict
communities taxonomic diversity response while functional diversity was
enhanced whenever the disturbance intensity. Although consistent, the
recovery of community composition, diversity and redundancy remained
unachieved after 30 years. This acknowledged the need of decades-long
cycles with no disturbance to ensure a complete recovery, and questioned
tropical forest community resilience after repeated disturbance.
}

@@

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna,
regulate the local and regional climates, the carbon, water and nutrient
cycles, and ensure cultural and human well-being. The growing demand in
forests products together with current global changes increases the
pressure on remaining natural forests \citep{Morales-Hidalgo2015} and
threatens the maintenance and dynamics in space and time of communities
structure, composition and functioning \citep{Anderson-Teixeira2013}.

In tropical forests, ecological communities are regularly re-shaped by
natural disturbance events changing both the abiotic environment,
through the fluxes of light, heat and water \citep{Goulamoussene2017},
and the biotic interactions such as competition among species
\citep{Chesson2000, Herault2018}. One of the cornerstone of tropical
forest ecology is to understand the processes and drivers of ecosystems
response to disturbance \citep{Chazdon2003a}. For now, this has been
largely studied through forest structural parameters such as aboveground
biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016} that are rapid and convenient to
measure. These structural parameters have been sucessfully modeled,
giving important insights into the recovery of ecosystem processes and
services \citep{Herault2018}. However the response of forests diversity
and composition is still unclear, albeit it determines the productivity,
stability and functioning of ecosystems \citep{Tilman2014, Liang2016}.
In the short-term, moderate disturbance may lead to positive impacts on
communities diversity, an idea formalized by the intermediate
disturbance hypothesis (IDH) stating a maximized species diversity when
disturbance intensity is not too high \citep{Molino2001, Kariuki2006a}.

Validations of the IDH though remain scarce in the long-term and mainly
rely on the analysis of taxonomic richness \citep{Molino2001}. Taxonomic
richness alone, though, gives limited or misleading information on
forests recovery and functioning \citep{Chaudhary2016}. More
ecological-meaningful analysis would couple richness with (i) evenness
that would reveal underlying ecological processes and (ii) composition
that is crucial for conservation issues
\citep{Lavorel2002, Bellwood2006}. Furthermore, a functional approach
accounting for species biological attributes would directly link
communities diversity, composition and redundancy to ecosystem
functioning and to its environmental constraints
\citep{Violle2007b, Baraloto2012a}. In that respect, the functional
trait-based approach that focus on major traits related to species
ecology and mediate species performance in a given environment was
sucessfully adopted \citep{Diaz2005}. For instance, the functional
approach revealed in tropical rainforests the deterministic processes
entailing, after disturbance, a functional shift from a dominance of
``conservative'' slow-growing species dealing with scarce resources to
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Reich2014, Herault2011}. This shift is
translated into the trajectories of key functional traits related to
resource acquisition (leaf and stem traits) and life-history traits
(seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. Eventually
a complete overview of communities response to disturbance would
encompass the changes in functional redundancy, that quantifies the
amount of shared trait values among species \citep{Carmona2016}. The
high functional redundancy of hyperdiverse tropical forests
\citep{Bellwood2006} mitigates the impacts of species removal on
ecosystem functioning and determines the resilience of communities after
disturbance \citep{Elmqvist2003, Diaz2005}.

In this study, we monitored over 30 years the response of 75 ha of
neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of ecosystem above-ground biomass (AGB) loss. We made
use of a large functional traits database encompassing major leaf, stem
and life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
Specifically, we (i) elucidated community taxonomic and functional
recovery and the determinism of underlying processes, (ii) clarified the
validity of the IDH in the long term for tropical forest and its
translation into different trajectories in time, and (iii) questioned
community recovery time.

@@

The experiment is a network of twelve 6.25ha plots that underwent a
disturbance gradient of three logging, thinning and fuelwood cutting
treatments (Table \ref{tab:Tab1}) according to a randomized plot design
with three replicate blocks of four plots. The disturbance corresponds
to averages of 10 trees removed per hectare with a diameter at 1.3 m
height (DBH) above 50 cm for treatment 1 (T1), 32 trees/ha above 40 cm
DBH for treatment 2 (T2) and 40 trees/ha above 40 cm DBH for treatment 3
(T3). Treatments T2 and T3 besides included the thinning of trees by
poison girdling \citep{Schmitt1990}. The disturbance intensity was
measured as the percentage of aboveground biomass (\%AGB) lost between
the first inventory in 1984 and five years after disturbance
\citep{Piponiot2016} estimated with the BIOMASS R package
\citep{Biomass2018}.

\begin{table}

\caption{\label{tab:Tab1}Intervention table, summary of the disturbance intensity for the 4 plot treatments in Paracou.}
\centering
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\toprule
Treatment & Timber & Thinning & Fuelwood & \%AGB lost\\
\midrule
Control &  &  &  & 0\\
T1 & DBH $\geq$ 50 cm, commercial species, $\approx$ 10 trees/ha &  &  & $[12\%-33\%]$\\
T2 & DBH $\geq$ 50 cm, commercial species, $\approx$ 10 trees/ha & DBH $\geq$ 40 cm, non-valuable species, $\approx$ 30 trees/ha &  & $[33\%-56\%]$\\
T3 & DBH $\geq$ 50 cm, commercial species, $\approx$ 10 trees/ha & DBH $\geq$ 50 cm, non-valuable species, $\approx$ 15 trees/ha & 40 cm $\leq$ DBH $\leq$ 50 cm, non-valuable species, $\approx$ 15 trees/ha & $[35\%-56\%]$\\
\bottomrule
\end{tabu}
\end{table}

@@

See Supplementary Materials -Figure S1 and \citet{Aubry-Kientz2013} for
the detailed methodology.

@@

To counter taxonomic uncertainties due to the variability of botanical
identification levels (in space) and protocols (in time), the taxonomic
composition and diversity analysis were conducted at the genus level.
Taxonomic and functional trajectories of community composition were
followed in a two-dimensional NMDS ordination space. Two NMDS using
abundance-based (Bray-Curtis) dissimilarity measures were conducted to
map either taxonomic or functional composition, the later based on the 7
leaf, stem and life history traits (without seed mass classes).
Trajectories along time were reported through the euclidean distance
between the target inventories and the reference inventories in 1989,
\emph{i.e.} 2 years after disturbance. Univariate trajectories of the
leaf, stem and life-history traits were also visualized with the
community weighted means (CWM) \citep{Diaz2007}. Species seed mass
corresponded to 5 classes of increasing mass, seed mass trajectories
were therefore reported as the proportion of each class in the
inventories (Supplementary materials).

The taxonomic diversity was reported through species richness and
evenness, \emph{i.e} the Hill number translation of the Simpson index
\citep{Hill1973}. These indices belong to the set of HCDT or generalized
entropy, respectively corresponding to the 0 and 2 order of diversity
(q), recomended for diversity studies \citep{Marcon2015b}. The
functional diversity was reported using the functional richness and
functional evenness, \emph{i.e} Rao index of quadratic entropy which
combines species abundance distribution and average pairwise
dissimilarity based on species functional traits.

@@

The functional redundancy was measured as the overlap among species in
community functional space \citep{Carmona2016}. The samples of the trait
database were first mapped in a 2-dimensional plan with a PCA analysis.
Then, multivariate kernel density estimator associated with individual
trees returned species traits probability distribution (TDP). Species
TDP weighted by species abundance were eventually summed for each
community. Community functional redundancy was the sum of TDPs overlap,
expressed as the average number of species that could be removed from
without reducing the functional space (Supp. Mat. - Figure S1 for a more
comprehensive sheme).

@@

\caption{Plot trajectories in terms of flora composition (left panels \textbf{(a)} and \textbf{(c)}) and functional composition (right panels \textbf{(b)} and \textbf{(d)}) in a two-dimensional NMDS space. Lower panels (\textbf{(c)} and \textbf{(d)}) represent the euclidean distance to initial condition along the 30 sampled years. Colors are treatments: green (control), yellow (T1), orange (T2), red (T3) with shaded areas the credibility intervals.}\label{fig:NMDSplans}
\end{figure*}

@@

Maximum height at adult stage (\emph{Hmax}), leaf toughness
(\emph{L\_toughness}) and wood specific gravity (\emph{WSG}) first
decreased and then slightly increased but remained significantly lower
than their initial value (Figure \ref{fig:CWM}). On the other side, Bark
thickness (\emph{Bark\_thick}) and specific leaf area (\emph{SLA})
increased and while \emph{Bark\_thick} remained substantially high after
30 years, \emph{SLA} had almost recovered its initial value. For all
traits, the maximum difference to initial value was correlated to the
disturbance intensity (\(\rho_{spearman}^{L_{thickness}}=0.76\),
\(\rho_{spearman}^{L_{chloro}}=0.60\),
\(\rho_{spearman}^{L_{toughness}}=-0.53\),
\(\rho_{spearman}^{SLA}=0.93\), \(\rho_{spearman}^{WSG}=-0.75\),
\(\rho_{spearman}^{Bark-thickness}=0.71\),
\(\rho_{spearman}^{Hmax}=-0.40\)). The proportions of the three lightest
seed mass classes increased in all disturbed plots, and decreased after
30 years for the lightest class while it stabilized for the two other
(Supp. Mat. - Figure S2).

@@

\caption{Trajectories of the communities weighted means (CWM) over 30 years after disturbance of 4 leaf traits (Leaf thickness, \emph{L\_thickness}, chlorophyll content, \emph{L\_chloro}, toughness, \emph{L\_toughness} and specific area, \emph{SLA}), 2 stem traits (wood specific gravity, \emph{WSG}, and bark thickness, \emph{Bark-thick}) and one life history trait (Specific maximum height at adult stage, \emph{Hmax}). Colors are treatments: green (control), yellow (T1), orange (T2), red (T3) with shaded areas the credibility intervals.}\label{fig:CWM}
\end{figure*}

@@

\caption{Trajectories over 30 years of the difference with the 1989 inventory (2 years after disturbance) of community taxonomic \textbf{(a)} richness, \textbf{(b)}, taxonomic evenness, \textbf{(c)} functional richness, and \textbf{(d)} functional evenness. Colors are treatments: green (control), yellow (T1), orange (T2), red (T3) with shaded areas the credibility intervals }\label{fig:DivTaxo}
\end{figure*}

@@

\caption{Relationship between the initial \%AGB loss and community taxonmic richness \textbf{(a)}, taxonomic evenness \textbf{(b)}, functional richness \textbf{(c)},and functional evenness \textbf{(d)} at 10, 20 and 30 years after disturbance. Colors are treatments: green (control), yellow (T1), orange (T2), red (T3) with shaded areas the credibility intervals }\label{fig:IDHplot}
\end{figure*}

@@

\caption{Trajectories of the functional redundancy within the initial functional space over 30 years after disturbance. Colors are disturbance treatments: green (control), yellow (T1), orange (T2), red (T3) with shaded areas the credibility intervals.}(\#fig:RedFun_rest)
\end{figure}

@@

Conversely, disturbed communities followed functional trajectories
similar in terms of functional composition. As pre-disturbance surviving
trees mirror the initial community \citep{Herault2018}, changes in
functional composition relied upon the recruitment of species or
functional types that were infrequent or absent before disturbance.
Competitive pioneers became dominant in filling the environmental niches
of high availability of light, space and nutrient vacated by the
disturbance. The recruitment of pioneers changed community functional
composition in the same way for all disturbance intensity towards more
resource-acquisitive strategies, moving community functional composition
right along the first axis in Figure \ref{fig:NMDSplans}
\citep{Westoby1998, Wright2004, Reich2014}. Thereafter long-lived, more
resistant and shade-tolerant species excluded the first established
pioneers and started the recovery of pre-disturbance functional
composition, moving similarly community functional composition left
along the first axis and upward along the second axis in Figure
\ref{fig:NMDSplans}.

@@

The IDH predicted well the disturbance impact on community taxonomic
richness, enhanced until an intensity threshold (20-25\% AGB loss), and
to some extent on taxonomic evenness, somewhat decoupled from the
disturbance intensity as already observed in the Guiana Shield
\citep{Baraloto2012a} and in Bornean tropical forests
\citep{Cannon1998}. The disturbance intensity determined the balance in
the community between pre-disturbance surviving trees and those
recruited afterward. The pool of true pioneer species specifically
recruited after disturbance is restricted in the Guiana Shield to a few
common genera (e.g.~Cecropia spp., Vismia spp.) \citep{Guitet2018}.
Below the intensity threshold the size of the surviving community
maintained the pre-disturbance high taxonomic richness while the
recruitment of pioneers, infrequent or absent before disturbance,
increased both community taxonomic richness and evenness. Beyond the
intensity threshold, the disturbance decreased the taxonomic richness of
surviving trees which was not offset by the enrichment of pioneers, so
that the overall community taxonomic richness decreased according to the
disturbance intensity \citep{Molino2001}. For community taxonomic
evenness the disturbance impact was similar but slighter, as the
evenness is less sensitive to the loss of rare species. Taxonomic
evenness rather represented the increasing dominance of pionneers that
balanced the usual hyper-dominance of tropical forest below the
intensity threshold, thus increasing community overall evenness up to
the intensity threshold beyond which pioneers became in turn highly
dominant and decreased the overall evenness \citep{Baraloto2012a}.

@@

Along time, taxonomic richness trajectories of all disturbed communities
first dropped similarly, following the species loss due to disturbance,
and then displayed a species gain depending on the disturbance
intensity. Up to an intensity threshold, the species gain was all the
more significant that the disturbance intensity increased, with the
establishment of long-lived pioneers enhancing community taxonomic
richness and evenness in the long term. These long-lived pionneers,
functionally quite different from the functional composition, entailed
as well a progressive and long-lasting increase of the functional
richness and evenness \citep{Denslow1980, Molino2001}. Beyond an
intensity threshold, though, a few short-lived pioneers occupied the
vacated environmental space and prevented the establishment of other
species. These short-lived pioneers were functionally very different
from the pre-disturbance community and entailed a rapid and significant
increase of functional richness en evenness. Already after 10 years,
though, short-lived pioneers started to decline and the functional
richness and evenness decreased. Likely this decrease will be followed
by the establishment of long-lasting pioneers, and by the time they
recruit we expect the taxonomic and functional trajectories to catch up
with those observed after intermediate disturbance \citep{Walker2009}.

@@

For 15 years the species loss during disturbance, determined by the
disturbance intensity, commensurately decreased the functional
redundancy within the pre-disturbance functional space. The redundancy
decrease was not compensated in the first place as the first recruited
pionneers were functionally different from the pre-disturbance
functional composition. Progressively though, first established species
were replaced by more competitive long-lived pionneers or
late-successional species resembling more the pre-disturbance functional
composition and restoring the functional redundancy. This replacement
was stochastic and followed the lottery recruitment rules, implying a
recruitment eased for the first recruited species but then increasingly
hampered by the emergence of interspecific competition
\citep{Busing2002}. Along time the recovery of infrequent species was
increasingly difficult, so that the time for the full recovery of the
functional redundancy, in some communities just initiated after 30
years, was difficult to estimate \citep{Elmqvist2003, Diaz2005}.

The long-term impact of disturbance on community functional redundancy
meant a lower resilience of the pre-disturbance communities, with higher
chances to see the persistence of disturbance-specific species at the
expense of late-successional ones \citep{Haddad2008}. Besides the
long-term recovery of infrequent species increases the risks to loose
cornerstone species, with unexpected ecological consequences
\citep{Jones1994, Chazdon2003a, Diaz2005, Gardner2007}. Apart from the
functional characteristics considered here, infrequent species might
indeed have unique functions in the ecosystem or be a key for some
fauna.
